Structure Learning for $\VAR(p)$ Models

### 2019-05-14 paper/paper.tex
Completed correctness proofs for Algorithm 1.

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

@@

\begin{definition}[Graph Theory Review]
  A \textit{graph} $\gcg = (\mathcal{V}, \mathcal{E})$ is simply a
  tuple of sets respectively called \textit{nodes} and \textit{edges}.
  Throughout this paper, we have in all cases
  $V = [n] \defeq \{1, 2, \ldots, n\}$.  We will also focus solely on
  \textit{directed} graphs, where the edges
  $\mathcal{E} \subseteq V \times V$ are \textit{ordered} pairs.

  A (directed) \textit{path} (of length $r$) between nodes $i, j$,
  denoted $\gcgpath{i}{j}$, is a sequence
  $a_0, a_1, \ldots, a_{r - 1}, a_r$ with $a_0 = i$ and $a_r = j$ such
  that $\forall\ 0 \le k \le r\ (a_k, a_{k + 1}) \in \mathcal{E}$.

@@

\begin{lemma}
  \label{lem:still_persistent}
  If $\gcg$ is a persistent DAG, any subgraph formed by eliminating nodes
  as well as all of the in and out edges thereof is still persistent.
\end{lemma}
\begin{proof}
  \hl{TODO -- is this needed?}
\end{proof}

@@

\begin{proposition}
  \label{prop:independent_confounders}
  If $\gcg$ is strongly causal we have distinct
  $k_1, k_2 \in \anc{i} \cap \anc{j}$ are confounders of $i, j$, then
  $k_1 \not\in \anc{k_2}$, $k_2 \not\in \anc{k_1}$ and
  $\not\exists \ell \in \anc{k_1} \cap \anc{k_2}$ confounding
  $k_1, k_2$.
\end{proposition}
\begin{proof}
  \hl{TODO}
\end{proof}

\begin{corollary}
  If $\gcg$ is strongly causal, then distinct confounders are
  uncorrelated.
\end{corollary}
\begin{proof}
  This follows directly by combining proposition
  \ref{prop:independent_confounders} with proposition \ref{prop:ancestor_properties}.  \hl{I need to update prop} \ref{prop:ancestor_properties} \hl{to note the strong non-correlation property.}
\end{proof}

@@

\begin{proposition}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is persistent DAG, then if $k$
  confounds $i, j$ we have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that $j \pwgc i$, the other being symmetric.  First
  note also that by proposition \ref{prop:sc_graph_common_anc} we
  cannot have $i \in \anc{j}$ or $j \in \anc{i}$ and therefore every
  $k \in \anc{i}\cap\anc{j}$ will be a confounder.

@@

  To this end, let $F(z)$ be an arbitrary but strictly causal linear,
  and equation \eqref{eqn:ancestor_expansion} to $x_i(t)$ and
  $\psi \defeq F(z)x_j(t)$:

@@

  Evidently, (\hl{This is clearly not immediately evident}) this is
  $0$ for each $F$ if and only if
  $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$ and
  $\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} = 0$, which by
  Collary \ref{cor:time_lag_cancellation} occurs if and only if
  $\tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})$, which is impossible
  since in general $\tau_0(\A_{jk}) < \infty$ and in a persistent
  graph $\tau_\infty(\A_{ik}) = \infty$.
\end{proof}

@@

\begin{theorem}[Pairwise Recovery]
  \label{thm:tree_recovery}
  If the Granger-Causality graph $\gcg$ for $x(t)$ is a strongly
  causal DAG then $\gcg$ can be inferred from pairwise causality
  tests alone.  The procedure can be carried out, assuming we have an
  oracle for pairwise causality, via algorithm (\ref{alg:pwgr}).

@@

    \BlankLine
    \caption{Pairwise Graph Recovery}
    \label{alg:pwgr}
    % \TitleOfAlgo{Pairwise Graph Recovery}
    \Input{Pairwise Granger-causality relations between $n$ processes whose joint Granger-causality relations are known to form a strongly causal DAG $\gcg$.}
    \Output{Edges $E = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of the graph $\gcg$.}
    \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
      $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\
      % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
      $k = 0$ \texttt{\# a counter used only for notation}}
    \BlankLine
    $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# potential edges}\\
    \While{$S_k \ne \emptyset$}{
      $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$  \texttt{\# parents of current layer}\\
      % $P_k \leftarrow P_{k - 1} \cup P_k$ \texttt{\# Expand parent set}\\
      $S_{k + 1} \leftarrow S_k \setminus P_k$ \texttt{\# remove from } $S$\\

      \;
      $C_k \leftarrow \{j \in S_{k + 1}\ |\ \forall s \in S_{k + 1}\ (s, j) \not\in W\}$   \texttt{\# potential children of }$P_k$\\

      \For{$r = 0, \ldots, k$}
      {
        \label{alg:inner_loop}
        $D_{kr} \leftarrow \{(i, j) \in P_{k - r} \times C_k\ |\ (i, j) \in W,\ \text{no } i \rightarrow j \text{ path in } E_k \}$\\%  \texttt{\# edges incident on }$C_k$\\
        $E_k \leftarrow E_k \cup D_{kr}$ \texttt{\# add edges to }$E_k$\\
        % $W_{k + 1} \leftarrow W_k \setminus D_{kr}$  \texttt{\# remove edges from consideration}\\
      }
      % $W_{k + 1} \leftarrow W_k \setminus \big(\bigcup_{r = 0}^k D_{kr}\big)$  \texttt{\# remove edges from consideration}\\

      \;
      $k \leftarrow k + 1$
    }
    $E \leftarrow E_{k + 1}$\\
    \Return{$E$}
  \end{algorithm}
\end{theorem}

We prove the theorem by establishing the correctness of algorithm (\ref{alg:pwgr}).  The idea is to iteratively ``peel away layers'' of nodes by removing the nodes that have no parents remaining, which always exist since the graph is acyclic.  The additional requirement of strong causality ensures firstly that all actual edges of $\gcg$ manifest in some way as pairwise relations (by proposition \ref{prop:pwgc_anc}), and secondly, as we will see, that parents of $C_k$ are in $\bigcup_{r = 0}^kP_{k - r}$ and can be determined by a pairwise test.

\begin{example}
  The set $W$ collects all of the edges that may be edges of $\gcg$.
  In reference to figure \ref{fig:example_fig3}, each of the solid
  black edges, as well as the dotted red edges will be included, but
  \textit{not} the bidirectional green dash-dotted edge.  That we exclude
  bidirectional edges is important since it will provide guarantees
  that $P_k$ is non-empty.  The groupings $P_0, \ldots, P_3$
  are also indicated in figure \ref{fig:example_fig3}.

  The algorithm proceeds first with the parentless nodes $1, 2$ on the
  initial iteration where the edge $(1, 3)$ is added to $E$.  On the
  next iteration, the edges $(3, 4), (2, 4), (3, 5)$ are added, and
  the false edges $(1, 4), (1, 5)$ is excluded due to the paths
  $1 \rightarrow 3 \rightarrow 4$ and $1 \rightarrow 3 \rightarrow 5$
  in E.  Finally, edge $(4, 6)$ are added, and the false
  $(1, 6), (3, 6), (2, 6)$ edges are similarly excluded due to the
  ordering of the inner loop.

  \begin{figure}
    \centering
    \caption{Example graph for Algorithm \ref{alg:pwgr}}
    \footnotesize{Black arrows indicate true parent-child
      relations.  Red dotted arrows indicate pairwise causality (due to
      ancestor relations), green dash-dotted arrows indicates
      bidirectional pairwise causality (due to the confounding node
      $1$).  Blue groupings indicate each $P_k$ in Algorithm
      \ref{alg:pwgr}.}
    \label{fig:example_fig3}

@@

  That we need to proceed backwards through $P_{k - r}$ as in the
  inner loop of line \ref{alg:inner_loop} can also be seen from this
  example, where if instead
  $D_k = \{(i, j) \in \Big(\bigcup_{r = 0}^k P_{k - r}\Big) \times
  C_k\ |\ i \pwgc j \}$ we would infer the false positive edge
  $1 \rightarrow 4$.  Moreover, the same example shows that if simply
  $D_k = \{(i, j) \in P_k \times C_k\ |\ i \pwgc j \}$, then the edge
  $1 \rightarrow 3$ would be missed.
\end{example}

\begin{lemma}
  \label{lem:inner_loop_lemma}
  At any given iteration, $k$, of Algorithm \ref{alg:pwgr}, if we have
  $e \in E$ if and only if $e \in \mathcal{E}$, then the inner loop (line \ref{alg:inner_loop})
  will add $(i, j) \in \big(\bigcup_{r = 0}^kP_{k - r}\big) \times C_k$ to $E$
  if $(i, j) \in \mathcal{E}$.  i.e. any real edge will be included.
\end{lemma}
\begin{proof}
  Let $j \in C_k$, $i \in \bigcup_{r = 0}^k P_{k - r}$, and assume
  $i \in \pa{j}$.

  First, suppose that there is no $i \rightarrow j$ path in $E$.  Then
  since $i \gc j$ we have $i \pwgc j$ and $j \npwgc i$ (i.e.
  $(i, j) \in W$) by Proposition \ref{prop:pwgc_anc} and Corollary
  \ref{cor:bidirectional_edge} so at least one of
  $\{D_{kr} \}_{r = 0}^k$ contains $(i, j)$ and hence $(i, j)$ will be
  added to $E$ upon the end of the iteration.

  Now, suppose instead that there is already an $i \rightarrow j$ path
  in $E$.  Given our assumption that $E$ contains edges if and only if
  they are edges of $\gcg$, then either we already have $(i, j) \in E$
  correctly, or there is more than one $i \rightarrow j$ path in
  $\gcg$, which is not possible since $\gcg$ is strongly causal.
\end{proof}

\begin{lemma}
  \label{lem:W_subset_E}
  $W \subseteq \mathcal{E}$
\end{lemma}
\begin{proof}
  By proposition \ref{prop:pwgc_anc} we have
  $j \in \pa{i} \Rightarrow j \pwgc i$ and by Corollary \ref{cor:bidirectional_edge}
  there are no bidirectional edges in $\gcg$.
\end{proof}

\hl{Must show that Algorithm terminates in $n$ steps.}
\begin{lemma}
  For any $i \in P_k$ we have $i \in C_{k - 1}$.  \hl{[Can't yet prove
    this part] Moreover, for each $i \in \bigcup_{k = 0}^n P_k$, if
    $(s, i) \in W$ then $(s, i) \in E$ upon the termination of
    Algorithm} \ref{alg:pwgr}.
\end{lemma}
\begin{proof}
  We prove the second part by induction, the first part being
  established along the way.  By construction $P_0$ has no parent
  nodes in $\gcg$ so the base case is immediate.  Fix now some
  $k \ge 1$, $i \in P_k$, and let $s$ be such that $(s, i) \in W$.
  For every $s$ such that $(s, i) \in W$ we have $s \not\in S_k$ since
  otherwise $i \not\in P_k$.  Therefore
  $s \in \bigcup_{r = 0}^{k - 1}P_{k - r}$, which implies
  $s \not\in S_k$, and finally that $i \in C_{k - 1}$.  Then by Lemma
  \ref{lem:W_subset_E} every edge in $\mathcal{E}$ is also present in
  $W$, and by Lemma \ref{lem:inner_loop_lemma} \hl{[We can't conclude
    here because of the strong conditions required on lemma}
    \ref{lem:inner_loop_lemma}]
\end{proof}

\begin{lemma}[Loop Invariant]
  \label{lem:loop_invariant}
  $\forall k \ge 0,\ e \in E_k \Rightarrow e \in \mathcal{E}$.
  \hl{Not sure if this is useful / the right approach.}
\end{lemma}
\begin{proof}
  We will proceed by induction.  Firstly, the base case
  $E_0 = \emptyset$ is trivial, so fix some $k \ge 1$ and consider an
  edge $(i, j) \in \big(\bigcup_{r = 0}^k P_{k - r} \big) \times C_k$
  such that $(i, j) \in W$ (if the edge is not in $W$ then it is not
  in $\mathcal{E}$ by Lemma \ref{lem:W_subset_E}).

  \hl{WIP}
\end{proof}

@@

\begin{lemma}
  $i \in P_k \iff d(i) = k$ and $j \in S_k \iff d(j) \ge k$.
\end{lemma}
\begin{proof}
  We proceed by induction.  The base case $j \in P_0 \iff d(j) = 0$ is
  by definition, and $j \in S_0 \iff d(j) \ge 0$ is trivial since
  $S_0 = [n]$.  So suppose that the lemma is true up to $k - 1$.

  For $i \in P_k$ we have by construction that there is no $j \in S_k$
  such that $(j, i) \in W$ and therefore no $j \in S_k$ such that
  $j \in \pa{i}$.  Therefore, any parent of $i$ can only be in
  $S_{k - r}$ for $r \ge 1$ and by the induction hypothesis which
  implies that $d(i) \le k$.  But if $d(i) < k$ then $i \in P_{k - r}$
  for some $r \ge 1$ (again by the induction hypothesis) thence
  $d(i) = k$ since $P_k \cap P_{k - r} = \emptyset$ by construction.

  Now for $s \in S_k$ we have by induction that and the fact that
  $s \in S_k \implies s \in S_{k - 1}$ by construction we have
  $d(s) \ge k - 1$, but again by induction (this time on $P_{k - 1}$)
  we have $d(s) \ne k$ since $S_k = S_{k - 1} \setminus P_{k - 1}$
  and therefore $d(s) \ge k$.

  Conversely, suppose $i \in [n]$ is such that $d(i) = k$.  Then
  $i \in S_{k - 1}$ by the hypothesis, but also $i \not\in P_{k - 1}$
  so then $i \in S_k = S_{k - 1} \setminus P_{k - 1}$.  Now, recalling
  the definition of $P_k$

@@

  if $s \in S_k$ is such that $s \pwgc i$ then by Proposition
  \ref{prop:ancestor_properties} $(a) s \in \anc{i}$ or
  $(b) \exists u \in \anc{i} \cap \anc{j}$ which is a confounder.  If
  we have $s \in \anc{i}$ then since $d(s) \ge k$ we must have
  $d(i) > k$, which is a contradiction so $s \not\in \anc{i}$.  On the
  other hand, if we have a confounding $u \in \anc{i} \cap \anc{s}$
  then we must have $d(u) < k$ (since $s \in \anc{i}$ and $d(i) = k$)
  which implies by induction that $u \in P_{k - r}$ for some $r \ge 1$,
  and therefore that $u \not\in ???$  \hl{This is wrong.}

  Now let $j \in S_k$, then $d(j) \ge k - 1$ since we
  also have $j \in S_{k - 1}$ (i.e. $S_k$ is a decreasing sequence of
  sets) \hl{Can't conclude unless $i \in P_k \iff d(i) = k$, in which
    case, we have $S_k = S_{k - 1} \setminus P_{k - 1}$ and therefore
    $d(j) \ge k$.}
\end{proof}

>>>>

\def\gc{\overset{\text{GC}}{\rightarrow}}  % Granger causality arrow
\def\ngc{\overset{\text{GC}}{\nrightarrow}}  % Negated Granger causality arrow
\def\pwgc{\overset{\text{PW}}{\rightarrow}}  % Pairwise Granger causality arrow
\def\npwgc{\overset{\text{PW}}{\nrightarrow}}  % Negated pairwise Granger causality arrow
\def\te{\overset{\mathcal{T}}{\rightarrow}}  % Transfer entropy arrow
\def\gcg{\mathcal{G}}  % Granger-causality graph
\def\gcge{\mathcal{E}}  % Graph edges
\def\VAR{\mathsf{VAR}}  % VAR(p) model
\def\B{\mathsf{B}}  % Filter B
\def\wtB{\widetilde{\B}}  % General filter B
\def\A{\mathsf{A}}  % Filter A
\def\H{\mathcal{H}}  % Hilbert space

@@

\begin{definition}[Graph Theory Review]
  A \textit{graph} $\gcg = (\mathcal{V}, \gcge)$ is simply a
  tuple of sets respectively called \textit{nodes} and \textit{edges}.
  Throughout this paper, we have in all cases
  $V = [n] \defeq \{1, 2, \ldots, n\}$.  We will also focus solely on
  \textit{directed} graphs, where the edges
  $\gcge \subseteq V \times V$ are \textit{ordered} pairs.

  A (directed) \textit{path} (of length $r$) between nodes $i, j$,
  denoted $\gcgpath{i}{j}$, is a sequence
  $a_0, a_1, \ldots, a_{r - 1}, a_r$ with $a_0 = i$ and $a_r = j$ such
  that $\forall\ 0 \le k \le r\ (a_k, a_{k + 1}) \in \gcge$.

@@

% \begin{lemma}
%   \label{lem:still_persistent}
%   If $\gcg$ is a persistent DAG, any subgraph formed by eliminating nodes
%   as well as all of the in and out edges thereof is still persistent.
% \end{lemma}
% \begin{proof}
%   \hl{TODO -- is this needed?}
% \end{proof}

@@

% \begin{proposition}
%   \label{prop:independent_confounders}
%   If $\gcg$ is strongly causal we have distinct
%   $k_1, k_2 \in \anc{i} \cap \anc{j}$ are confounders of $i, j$, then
%   $k_1 \not\in \anc{k_2}$, $k_2 \not\in \anc{k_1}$ and
%   $\not\exists \ell \in \anc{k_1} \cap \anc{k_2}$ confounding
%   $k_1, k_2$.
% \end{proposition}
% \begin{proof}
%   \hl{TODO}
% \end{proof}

% \begin{corollary}
%   If $\gcg$ is strongly causal, then distinct confounders are
%   uncorrelated.
% \end{corollary}
% \begin{proof}
%   This follows directly by combining proposition
%   \ref{prop:independent_confounders} with proposition \ref{prop:ancestor_properties}.  \hl{I need to update prop} \ref{prop:ancestor_properties} \hl{to note the strong non-correlation property.}
% \end{proof}

@@

\begin{proposition}
  \label{prop:persistence_converse}
  Suppose $\gcg$ is a strongly causal DAG and that $x(t)$ is persistent, then if $k$
  confounds $(i, j)$ we have $i \pwgc j$ and $j \pwgc i$.
\end{proposition}
\begin{proof}
  We will show that $j \pwgc i$, the other being symmetric.  First
  note also that by proposition \ref{prop:sc_graph_common_anc} we
  cannot have $i \in \anc{j}$ or $j \in \anc{i}$ and therefore every
  $k \in \anc{i}\cap\anc{j}$ will be a confounder.

@@

  To this end, let $F(z)$ be an arbitrary but strictly causal linear
  filter.  We apply equation \eqref{eqn:ancestor_expansion} to
  $x_i(t)$ and $\psi \defeq F(z)x_j(t)$:

@@

  Evidently, (\hl{This is clearly not immediately evident}) this is
  $0$ for each $F$ if and only if
  $\linE{F(z)\A_{jk}(z)v_k(t)}{\H_{t - 1}^{(i)}} = 0$ and
  $\inner{F(z)\A_{jk}(z)v_k(t)}{\A_{ik}(z)v_k(t)} = 0$, which by
  Collary \ref{cor:time_lag_cancellation} occurs if and only if
  $\tau_0(\A_{jk}) \ge \tau_\infty(\A_{ik})$, which is impossible
  since in general $\tau_0(\A_{jk}) < \infty$ and for a persistent system
  $\tau_\infty(\A_{ik}) = \infty$.
\end{proof}

@@

\begin{theorem}[Pairwise Recovery]
  \label{thm:tree_recovery}
  If the Granger-causality graph $\gcg$ for persistent process $x(t)$
  is a strongly causal DAG then $\gcg$ can be inferred from pairwise
  causality tests alone.  The procedure can be carried out, assuming
  we have an oracle for pairwise causality, via algorithm
  (\ref{alg:pwgr}).

@@

    \BlankLine
    \caption{Pairwise Graph Recovery}
    \label{alg:pwgr}
    % \TitleOfAlgo{Pairwise Graph Recovery}
    \Input{Pairwise Granger-causality relations between a persistent
      processes of dimension $n$ whose joint Granger-causality
      relations are known to form a strongly causal DAG $\gcg$.}
    \Output{Edges $\gcge = \{(i, j) \in [n] \times [n]\ |\ i \gc j \}$ of
      the graph $\gcg$.}
    \Initialize{$S_0 = [n]$  \texttt{\# unprocessed nodes}\\
      $E_0 = \emptyset$  \texttt{\# edges of }$\gcg$\\
      % $P_0 = \emptyset$  \texttt{\# layer by layer driving nodes}\\
      $k = 1$ \texttt{\# a counter used only for notation}}
    \BlankLine
    $W \leftarrow \{(i, j)\ |\ i \pwgc j, j \npwgc i\}$  \texttt{\# candidate edges}\\
    $P_0 \leftarrow \{i \in S_0\ |\ \forall s \in S_0\ (s, i) \not\in W\}$  \texttt{\# parentless nodes}\\
    \While{$S_{k - 1} \ne \emptyset$}{
      $S_k \leftarrow S_{k - 1} \setminus P_{k - 1}$ \texttt{\# remove nodes with depth }$k - 1$\\
      $P_k \leftarrow \{i \in S_k\ |\ \forall s \in S_k\ (s, i) \not\in W\}$   \texttt{\# candidate children of }$P_{k - 1}$\\
      \;

      $D_{k0} \leftarrow \emptyset$\\
      \For{$r = 1, \ldots, k$} 
      {
        $Q \leftarrow E_{k - 1} \cup \big(\bigcup_{\ell = 0}^{r - 1} D_{k\ell}\big)$ \texttt{\# currently known edges}\\
        $D_{kr} \leftarrow \{(i, j) \in P_{k - r} \times P_k\ |\ (i, j) \in W,\ \text{no } \gcgpath{i}{j} \text{ path in } Q\}$
        % $E_k \leftarrow E_{k - 1} \cup D_{kr}$ \texttt{\# add edges to }$E_k$ \label{alg:inner_loop_end}\\
        % $W_{k + 1} \leftarrow W_k \setminus D_{kr}$  \texttt{\# remove edges from consideration}\\
      }
      $E_k \leftarrow E_{k - 1} \cup \big(\bigcup_{r = 1}^k D_{kr}\big)$ \texttt{\# update } $E_k$ \texttt{ with new edges}\\
      % \label{alg_line:inner_loop_end}\\
      % $W_{k + 1} \leftarrow W_k \setminus \big(\bigcup_{r = 0}^k D_{kr}\big)$  \texttt{\# remove edges from consideration}\\
      \;
      $k \leftarrow k + 1$
    }
    \Return{$E_{k - 1}$}
  \end{algorithm}
\end{theorem}

We will shortly prove the theorem by establishing the correctness of
algorithm (\ref{alg:pwgr}).  The idea is to iteratively ``peel away
layers'' of nodes by removing the nodes that have no parents
remaining, which always exist since the graph is acyclic.  The
requirement of strong causality ensures that all actual edges of
$\gcg$ manifest in some way as pairwise relations (by proposition
\ref{prop:pwgc_anc}), and the requirement of persistence allows
confounding to be eliminated by removing bidirectional edges.  Without
persistence, each confounded pair would give rise to $4$ possible
pairwise topologies consistent with $\gcg$, one for each type of
pairwise edge (no edge, unidirectional, bidirectional).

\begin{example}
  The set $W$ collects ancestor relations in $\gcg$ (see Lemma
  \ref{lem:W_subset_E}).  In reference to figure
  \ref{fig:example_fig3}, each of the solid black edges, as well as
  the dotted red edges will be included in $W$, but \textit{not} the
  bidirectional green dash-dotted edges, which we are able to exclude
  by appealing to Corollary \ref{cor:bidirectional_edge}.  The
  groupings $P_0, \ldots, P_3$ are also indicated in figure
  \ref{fig:example_fig3}.

  The algorithm proceeds first with the parentless nodes $1, 2$ on the
  initial iteration where the edge $(1, 3)$ is added to $E$.  On the
  next iteration, the edges $(3, 4), (2, 4), (3, 5)$ are added, and
  the false edges $(1, 4), (1, 5)$ is excluded due to the paths
  $1 \rightarrow 3 \rightarrow 4$ and $1 \rightarrow 3 \rightarrow 5$
  already present.  Finally, edge $(4, 6)$ is added, and the false
  $(1, 6), (3, 6), (2, 6)$ edges are similarly excluded due to the
  ordering of the inner loop.

  \begin{figure}
    \centering
    \caption{Example graph for Algorithm \ref{alg:pwgr}}
    \footnotesize{Black arrows indicate true parent-child
      relations.  Red dotted arrows indicate pairwise causality (due to
      non-parent relations), green dash-dotted arrows indicates
      bidirectional pairwise causality (due to the confounding node
      $1$).  Blue groupings indicate each $P_k$ in Algorithm
      \ref{alg:pwgr}.}
    \label{fig:example_fig3}

@@

  That we need to proceed backwards through $P_{k - r}$ as in the
  inner loop of on $r$ can also be seen from this example, where if
  instead we simply added the set

  \begin{equation*}
    D_k' = \{(i, j) \in \Big(\bigcup_{r = 1}^k P_{k - r}\Big) \times P_k\ |\ i \pwgc j \}
  \end{equation*}

  to $E_k$ then we would infer the false positive edge
  $1 \rightarrow 4$.  Moreover, the same example shows that simply
  using the set

  \begin{equation*}
    D_k'' = \{(i, j) \in P_{k - 1} \times P_k\ |\ i \pwgc j \}  ,
  \end{equation*}

  causes the edge $1 \rightarrow 3$ to be missed.
\end{example}

% \begin{lemma}
%   \label{lem:inner_loop_lemma}
%   At any given iteration, $k$, of Algorithm \ref{alg:pwgr}, if we have
%   $e \in E$ if and only if $e \in \gcge$, then the inner loop (line \ref{alg:inner_loop})
%   will add $(i, j) \in \big(\bigcup_{r = 0}^kP_{k - r}\big) \times C_k$ to $E$
%   if $(i, j) \in \gcge$.  i.e. any real edge will be included.
% \end{lemma}
% \begin{proof}
%   Let $j \in C_k$, $i \in \bigcup_{r = 0}^k P_{k - r}$, and assume
%   $i \in \pa{j}$.

%   First, suppose that there is no $i \rightarrow j$ path in $E$.  Then
%   since $i \gc j$ we have $i \pwgc j$ and $j \npwgc i$ (i.e.
%   $(i, j) \in W$) by Proposition \ref{prop:pwgc_anc} and Corollary
%   \ref{cor:bidirectional_edge} so at least one of
%   $\{D_{kr} \}_{r = 0}^k$ contains $(i, j)$ and hence $(i, j)$ will be
%   added to $E$ upon the end of the iteration.

%   Now, suppose instead that there is already an $i \rightarrow j$ path
%   in $E$.  Given our assumption that $E$ contains edges if and only if
%   they are edges of $\gcg$, then either we already have $(i, j) \in E$
%   correctly, or there is more than one $i \rightarrow j$ path in
%   $\gcg$, which is not possible since $\gcg$ is strongly causal.
% \end{proof}

% \begin{lemma}
%   \label{lem:W_subset_E}
%   $W \subseteq \gcge$
% \end{lemma}
% \begin{proof}
%   By proposition \ref{prop:pwgc_anc} we have
%   $j \in \pa{i} \Rightarrow j \pwgc i$ and by Corollary \ref{cor:bidirectional_edge}
%   there are no bidirectional edges in $\gcg$.
% \end{proof}

% \hl{Must show that Algorithm terminates in $n$ steps.}
% \begin{lemma}
%   For any $i \in P_k$ we have $i \in C_{k - 1}$.  \hl{[Can't yet prove
%     this part] Moreover, for each $i \in \bigcup_{k = 0}^n P_k$, if
%     $(s, i) \in W$ then $(s, i) \in E$ upon the termination of
%     Algorithm} \ref{alg:pwgr}.
% \end{lemma}
% \begin{proof}
%   We prove the second part by induction, the first part being
%   established along the way.  By construction $P_0$ has no parent
%   nodes in $\gcg$ so the base case is immediate.  Fix now some
%   $k \ge 1$, $i \in P_k$, and let $s$ be such that $(s, i) \in W$.
%   For every $s$ such that $(s, i) \in W$ we have $s \not\in S_k$ since
%   otherwise $i \not\in P_k$.  Therefore
%   $s \in \bigcup_{r = 0}^{k - 1}P_{k - r}$, which implies
%   $s \not\in S_k$, and finally that $i \in C_{k - 1}$.  Then by Lemma
%   \ref{lem:W_subset_E} every edge in $\gcge$ is also present in
%   $W$, and by Lemma \ref{lem:inner_loop_lemma} \hl{[We can't conclude
%     here because of the strong conditions required on lemma}
%     \ref{lem:inner_loop_lemma}]
% \end{proof}

% \begin{lemma}[Loop Invariant]
%   \label{lem:loop_invariant}
%   $\forall k \ge 0,\ e \in E_k \Rightarrow e \in \gcge$.
%   \hl{Not sure if this is useful / the right approach.}
% \end{lemma}
% \begin{proof}
%   We will proceed by induction.  Firstly, the base case
%   $E_0 = \emptyset$ is trivial, so fix some $k \ge 1$ and consider an
%   edge $(i, j) \in \big(\bigcup_{r = 0}^k P_{k - r} \big) \times C_k$
%   such that $(i, j) \in W$ (if the edge is not in $W$ then it is not
%   in $\gcge$ by Lemma \ref{lem:W_subset_E}).

%   \hl{WIP}
% \end{proof}

Our proof proceeds in 5 steps stated formally as lemmas.  Firstly, we
characterize the sets $W$ and $P_k$.  Then we establish a correctness
result for the inner loop on $r$, a correctness result for the outer
loop on $k$, and finally that the algorithm terminates in a finite
number of steps.

\begin{lemma}[$W$ Represents Ancestor Relations]
  \label{lem:W_subset_E}
  In Algorithm \ref{alg:pwgr} we have
  $(i, j) \in W$ if and only if $i \in \anc{j}$.  In particular,
  $W \subseteq \gcge$.
\end{lemma}
\begin{proof}
  Let $j \in [n]$ and suppose that $i \in \anc{j}$.  Then $i \pwgc j$
  by Proposition \ref{prop:pwgc_anc}.  Proposition
  \ref{prop:sc_graph_common_anc} ensures that $(i, j)$ are not
  confounded and Corollary \ref{cor:parent_corollary} that
  $j \not\in \anc{i}$ so $j \npwgc i$ and thence by Proposition
  \ref{prop:ancestor_properties} $(i, j) \in W$.

  Convsersely, suppose $(i, j) \in W$.  Then since $j \npwgc i$
  Proposition \ref{prop:persistence_converse} ensures that $(j, i)$
  are not confounded and so by Proposition \ref{prop:ancestor_properties}
  we must have $i \in \anc{j}$.
\end{proof}

@@

\begin{lemma}[Depth Characterization of $P_k$]
  \label{lem:depth_lemma}
  $i \in P_k \iff d(i) = k$ and $j \in S_k \iff d(j) \ge k$.
\end{lemma}
\begin{proof}
  We proceed by induction, noting that $P_0$ is non-empty since $\gcg$
  is acyclic and therefore $\gcg$ contains nodes without parents.  The
  base case $i \in P_0 \iff d(i) = 0$ is by definition, and
  $j \in S_0 \iff d(j) \ge 0$ is trivial since $S_0 = [n]$.  So
  suppose that the lemma is true up to $k - 1$.

  ($i \in P_k \implies d(i) = k$): Let $i \in P_k$.  Suppose that
  $d(i) \ge k + 1$, then $\exists j \in \pa{i}$ such that
  $j \not\in \cup_{r \ge 1}P_{k - r}$ (otherwise $d(i) \le k$), this
  implies that $j \in S_k$ with $(j, i) \in W$ (by Lemma
  \ref{lem:W_subset_E}) which is not possible due to the construction of
  $P_k$ and therefore $d(i) \le k$.  Moreover,
  $P_k \subseteq S_k \subseteq S_{k - 1}$ implies that
  $d(i) \ge k - 1$ by the induction hypothesis, but if $d(i) = k - 1$
  then $i \in P_{k - 1}$ again by induction which is impossible since
  $i \in P_k$ and therefore $d(i) = k$.

  ($s \in S_k \implies d(s) \ge k$): Let
  $s \in S_k \subseteq S_{k - 1}$.  We have by induction that
  $d(s) \ge k - 1$, but again by induction (this time on $P_{k - 1}$)
  we have $d(s) \ne k - 1$ since $S_k = S_{k - 1} \setminus P_{k - 1}$
  and therefore $d(s) \ge k$.

  ($d(i) = k \implies i \in P_k$): Suppose $i \in [n]$ is such that
  $d(i) = k$.  Then $i \in S_{k - 1}$ by the hypothesis, but also
  $i \not\in P_{k - 1}$ so then
  $i \in S_k = S_{k - 1} \setminus P_{k - 1}$ and thus $d(i) \ge k$.
  Now, recalling the definition of $P_k$

@@

  if $s \in S_k$ is such that $(s, i) \in W$ then $s \pwgc i$ and
  $i \npwgc s$ so that by persistence and Proposition
  \ref{prop:persistence_converse} there cannot be a counfounder of
  $(s, i)$ (otherwise $i \pwgc s$) so then by Proposition
  \ref{prop:ancestor_properties} we have $s \in \anc{i}$.  We have
  shown that $s \in S_k \implies d(s) \ge k$ and so we must have
  $d(i) > k$, a contradiction, thence $s \not\in \anc{i}$,
  $s \npwgc i$, $(s, i) \not\in W$ and $i \in P_k$.

  ($d(j) \ge k \implies j \in S_k$): Let $j \in [n]$ such that
  $d(j) \ge k$, then by induction we have $j \in S_{k - 1}$.  This
  implies by the construction of $S_k$ that $j \not\in S_k$ only if
  $j \in P_{k - 1}$, but we have shown that this only occurs when
  $d(j) = k - 1$, but $d(j) > k - 1$ so $j \in S_k$.
\end{proof}
